---
title: "Getting Started with figsr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with figsr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction to FIGS

**figsr** implements Fast Interpretable Greedy-Tree Sums (FIGS) (*Tan et al., PNAS
2023*). FIGS fits a *sum* of shallow decision trees,
$\hat{f}(x) = \sum_k \hat{f}_k(x)$, one split at a time.

At every step the algorithm compares a single global pool of candidates: opening
a new tree on the full sample, and splitting each leaf of every tree already
grown. The candidate with the largest reduction in the residual sum of squares
wins, and all residuals are recomputed against the whole sum before the next
step. That is the whole idea: *grow a new tree or deepen an existing one,
whichever helps most*.

The pay-off is that additive structure is modeled additively. A single CART
tree has to repeat a subtree for every combination of two independent effects;
FIGS puts each effect in its own small tree.

## A regression fit

```{r example}
library(figsr)

set.seed(42)
df <- data.frame(x1 = rnorm(300), x2 = rnorm(300), x3 = rnorm(300))
df$y <- 3 * (df$x1 > 0) + 2 * (df$x2 > 0.5) - 1.5 * (df$x3 < -0.2) +
  rnorm(300, sd = 0.3)

fit <- figs(y ~ x1 + x2 + x3, data = df, max_splits = 6)
fit
```

The three effects were generated independently, so we expect FIGS to recover
them as separate trees rather than as one deep tree.

```{r rules}
summary(fit)
```

Each leaf holds that tree's *contribution* to the prediction; a prediction is
the sum of one leaf per tree.

```{r predict}
preds <- predict(fit, new_data = df)
head(preds)

cor(preds$.pred, df$y)
```

## Variable importance

`figsr_importance()` adds up, for each predictor, the residual sum-of-squares
reduction of every split it carries.

```{r importance}
figsr_importance(fit)
```

## Visualizing the tree sum

`plot()` draws every tree in the sum. Three styles are available:
`"scientific"` (the default), `"modern"` and `"classic"`.

```{r plot, fig.alt = "The trees of the fitted FIGS model, drawn side by side."}
plot(fit)
```

## Two-class classification

A factor outcome with two levels switches the model to classification. The
engine still fits squared error, on the 0/1 encoding of the outcome, so the sum
of the leaf values estimates the probability of the second level directly.

```{r classification}
set.seed(7)
dfc <- data.frame(x1 = rnorm(300), x2 = rnorm(300))
score <- 1.5 * dfc$x1 + dfc$x2
dfc$y <- factor(ifelse(score + rnorm(300, sd = 0.5) > 0, "yes", "no"))

fit_c <- figs(y ~ x1 + x2, data = dfc, max_splits = 6)

head(predict(fit_c, new_data = dfc))
head(predict(fit_c, new_data = dfc, type = "prob"))
```

## Missing values

By default `figs()` drops any row with a missing predictor, as most
model-frame based fitting functions in R do. With many predictors that can
discard a large share of a small data set. Setting `na_method = "mia"` keeps
every row instead: at each candidate split the search also tries sending the
missing values to either side, and splitting on whether the value is missing
at all, and keeps whichever helps most. When missingness carries information,
the tree can say so in a rule.

```{r missing}
set.seed(7)
n <- 200
dm <- data.frame(x = rnorm(n), z = rnorm(n))
dm$x[sample.int(n, 60)] <- NA
dm$y <- 4 * is.na(dm$x) + 2 * (dm$z > 0) + rnorm(n, sd = 0.3)

fit_mia <- figs(y ~ x + z, data = dm, max_splits = 3, na_method = "mia")
summary(fit_mia)
figsr_importance(fit_mia)
```

A split on missingness is listed as `missing(x)` in the importance table,
because it never reads the value of `x`. If you would rather fill the gaps
before fitting, use a `recipes` imputation step such as
`recipes::step_impute_knn()` inside a workflow; `figsr` does not impute.

## Use inside tidymodels

`figs_tree()` registers FIGS with `parsnip`, so the model can be used anywhere a
`parsnip` specification is accepted, and `max_splits`, `max_trees` and `min_n`
can be tuned with `dials` and `tune`.

```{r parsnip}
library(parsnip)

spec <- figs_tree(max_splits = 6, min_n = 5) |>
  set_engine("figsr") |>
  set_mode("regression")

wf_fit <- fit(spec, y ~ x1 + x2 + x3, data = df)
head(predict(wf_fit, new_data = df))
```

`na_method` is an engine argument, so it is set in `set_engine()`:
`set_engine("figsr", na_method = "mia")`.

## Bootstrap ensembling

`bagging_figs()` fits several FIGS models on bootstrap resamples and averages
them. It trades the readable rule set for stability on noisy data.

```{r bagging}
bag <- bagging_figs(y ~ x1 + x2 + x3, data = df, n_estimators = 5, max_splits = 6)
head(predict(bag, new_data = df))
```

## Limitations

* Classification is limited to two classes.
* Case weights are not supported.
* Factor predictors with more than 10 levels present in a node are skipped,
  because the split search enumerates subsets of levels.
* Missing predictor values are handled by `na_method = "mia"`; there are no
  surrogate splits, and a predictor that was complete in training cannot be
  missing at prediction time.
