---
title: "Missing Values in figsr: A Worked Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing Values in figsr: A Worked Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(figsr)
```

The introductory vignette (`vignette("figsr-intro")`) covers `na_method =
"mia"` in a few lines on a toy data set. This vignette goes deeper, in two
steps: a synthetic case built so that missingness is the *entire* signal,
and a real clinical data set where missingness carries a real but modest
signal, mixed in with everything else a tree can learn.

## When missingness is the whole story

Some measurements are only taken when a clinician already suspects a
problem. An expensive lab test ordered only for patients who look sick is
not missing at random: the decision to order it is itself informative,
often more informative than the number that comes back.

```{r synthetic}
set.seed(7)
n <- 2000

age         <- pmin(pmax(rnorm(n, 55, 12), 30), 85)
cholesterol <- pmin(pmax(rnorm(n, 200, 38), 120), 350)

# The test is ordered more often for older, higher-cholesterol patients, but
# rarely overall -- an expensive, targeted test.
suspicion <- plogis(-4 + 0.05 * (age - 55) + 0.02 * (cholesterol - 200))
ordered   <- rbinom(n, 1, suspicion)

# Ordering the test is itself the dominant signal for the outcome.
risk_score <- 0.01 * (age - 55) + 0.005 * (cholesterol - 200) + 6 * ordered + rnorm(n)
bnp <- ifelse(ordered == 1, rnorm(n, mean = 400 + 2 * risk_score, sd = 60), NA)

screening <- data.frame(age, cholesterol, bnp, risk_score)

fit_screening <- figs(risk_score ~ ., data = screening, max_splits = 2, na_method = "mia")
summary(fit_screening)
figsr_importance(fit_screening)
```

The root of the first tree is a pure split on missingness -- `bnp` was
ordered, or it was not -- and that one rule carries 95% of the total gain.
`figsr_importance()` lists it on its own row, `missing(bnp)`, because the
split never reads the value of `bnp` at all.

## A real data set: the Pima Indians Diabetes Database

`pima_diabetes` ships with `figsr`; loading the package makes it available,
the same way `iris` is available after `library(datasets)`. It is 768
patient records with eight clinical predictors and a binary outcome
(diabetes within five years), one of the most-used data sets on Kaggle
(`kaggle.com/datasets/uciml/pima-indians-diabetes-database`), originally
distributed through the UCI Machine Learning Repository.

```{r load}
str(pima_diabetes)
```

Five columns record measurements that cannot physiologically be zero:
glucose, blood pressure, skin thickness, insulin and BMI. In the raw data a
`0` in any of them is a missing value recorded as if it were data -- a
well-documented quirk of this particular data set, not a `figsr` concern.
That recoding has already been done in `pima_diabetes`; see `?pima_diabetes`
for the exact preprocessing.

```{r missingness}
predictors_with_gaps <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI")
missing_rate <- vapply(pima_diabetes[predictors_with_gaps], function(x) mean(is.na(x)), numeric(1))
knitr::kable(
  data.frame(predictor = names(missing_rate),
             missing_pct = sprintf("%.1f%%", 100 * missing_rate)),
  row.names = FALSE
)
```

Insulin is missing in nearly half the records, and skin thickness in almost
a third. Only 51% of the rows are complete on all five columns at once --
`na_method = "omit"`, the default, would fit on barely half the data and
would be unable to predict for most new patients either.

```{r fit}
set.seed(42)
train_idx <- sample(nrow(pima_diabetes), floor(0.75 * nrow(pima_diabetes)))
train <- pima_diabetes[train_idx, ]
test  <- pima_diabetes[-train_idx, ]

fit_omit <- figs(Outcome ~ ., data = train, max_splits = 10, mode = "classification")
fit_mia  <- figs(Outcome ~ ., data = train, max_splits = 10, mode = "classification",
                 na_method = "mia")

cat("training rows used, omit:", length(fit_omit$fitted_values), "of", nrow(train), "\n")
cat("training rows used, mia :", length(fit_mia$fitted_values),  "of", nrow(train), "\n")

test_complete <- complete.cases(test[, predictors_with_gaps])
cat("test patients omit can score at all:", sum(test_complete), "of", nrow(test), "\n")
cat("test patients mia can score        :", nrow(test), "of", nrow(test), "\n")
```

`na_method = "mia"` fits on every training row and scores every test
patient, including the ones with an incomplete workup. `na_method = "omit"`
(the default) trains on about half the data and, at prediction time,
raises an error for a patient missing a predictor the tree actually split
on -- it can only be scored on the subset of test patients who happen to
have every measurement. This is the practical cost of the default: it is
not that it predicts worse, it is that it often cannot predict at all.

```{r accuracy}
accuracy <- function(predicted, observed) mean(predicted == observed)

pred_omit <- predict(fit_omit, test[test_complete, ], type = "class")$.pred_class
pred_mia  <- predict(fit_mia, test, type = "class")$.pred_class

cat("accuracy, omit, on the", sum(test_complete), "patients it can score:",
    round(accuracy(pred_omit, test$Outcome[test_complete]), 3), "\n")
cat("accuracy, mia, on those same", sum(test_complete), "patients        :",
    round(accuracy(pred_mia[test_complete], test$Outcome[test_complete]), 3), "\n")
cat("accuracy, mia, on all", nrow(test), "test patients                  :",
    round(accuracy(pred_mia, test$Outcome), 3), "\n")
```

On the subset both models can score, accuracy is close either way -- with a
single train/test split of 100-odd patients that difference is well within
sampling noise, and `mia` is not being sold here as more accurate on data it
was never asked to handle in the first place. The real result is coverage:
`mia` scores every patient in the test set, `omit` scores only those with a
complete workup.

```{r rules}
summary(fit_mia)
```

```{r importance}
figsr_importance(fit_mia)
```

`Glucose`, `BMI`, `SkinThickness` and `BloodPressure` each pick up an
`OR ... IS MISSING` clause somewhere in the sum: the tree folds those
missing values into whichever side of a split fits best, rather than giving
missingness its own rule the way `bnp` got one in the synthetic example
above. `Insulin` -- missing in nearly half the records, more than any other
column here -- carries no importance at all in this fit: `Glucose`, `Age`
and `BMI` already explain most of the outcome, and the greedy search never
needs it. This is the ordinary case for real data: missingness is one
signal among several, not the whole story, and a tree that is free to use
it will use it exactly as much as it helps, on whichever columns it helps,
and no more.

## When to impute instead

`figsr` does not impute. If a predictor is missing because of a process you
want to model away rather than learn from -- a sensor dropout, a scanning
error -- an imputation step such as `recipes::step_impute_knn()` inside a
`tidymodels` workflow is a better fit than `na_method = "mia"`. Use `"mia"`
when the fact that a value is missing might itself be informative, as it
plausibly is for both examples in this vignette.

## References

Smith, J. W., Everhart, J. E., Dickson, W. C., Knowler, W. C., and Johannes,
R. S. (1988). Using the ADAP learning algorithm to forecast the onset of
diabetes mellitus. *Proceedings of the Symposium on Computer Applications
and Medical Care*, 261-265.

Twala, B. E. T. H., Jones, M. C., and Hand, D. J. (2008). Good methods for
coping with missing data in decision trees. *Pattern Recognition Letters*,
29(7), 950-956.
